# Robust Efficiency Simulation {#sec-robust-efficiency}

The simulation does the following:

1.  Loop over contamination rates from 0 % to 30 %.\
2.  For each rate, repeat 100 Monte-Carlo draws:\
3.  simulate a clean exponential-decay sample,\
4.  add large noise to the chosen fraction of points,\
5.  fit both NLS and MM, and record each method’s mean-squared error.\
6.  Average the errors across repeats for every rate.\
7.  Plot the averages on a log-10 scale so the two curves (NLS in grey, MM in black) reveal how error grows with contamination.

```{r}
#| label: fig-efficiency-sim
#| fig-cap: "Mean-squared error for NLS and MM estimates as contamination increases"
#| echo: true
#| eval: false
library(robustbase)
library(ggplot2)
set.seed(123)
contam   <- seq(0, 0.30, 0.05)        # bad points
n_rep    <- 100                       # Monte-Carlo repeats
bounds_l <- c(a = 0,  b = 0, c = -5)
bounds_u <- c(a = 20, b = 2, c =  5)
mse_tbl <- data.frame()

for (alpha in contam) {
  mse_nls <- mse_mm <- numeric(n_rep)

  for (i in seq_len(n_rep)) {
    x <- runif(50, 0, 5)
    y <- 10 * exp(-0.5 * x) + 1 + rnorm(50, 0, 0.3)

    k <- floor(50 * alpha)            
    if (k > 0) {
      idx <- sample(50, k)
      y[idx] <- y[idx] + rnorm(k, 0, 5)
    }

    dat <- data.frame(x, y)

    ## NLS  --------------------------------------------------------------
    nls_fit <- try(
      nls(y ~ a * exp(-b * x) + c,
          dat,
          start     = list(a = 8, b = 0.4, c = 0.5),
          algorithm = "port",
          lower     = bounds_l,
          upper     = bounds_u),
      silent = TRUE)

    if (!inherits(nls_fit, "try-error"))
      mse_nls[i] <- mean((coef(nls_fit) - c(10, 0.5, 1)) ^ 2)
    else
      mse_nls[i] <- NA_real_

    ## MM  --------------------------------------------------------------
    mm_fit <- suppressWarnings(
      nlrob(y ~ a * exp(-b * x) + c,
            dat,
            lower  = bounds_l,
            upper  = bounds_u,
            method = "MM"))

    mse_mm[i] <- mean((coef(mm_fit) - c(10, 0.5, 1)) ^ 2)
  }

  mse_tbl <- rbind(mse_tbl,
                   data.frame(Contam = alpha, Method = "NLS",
                              MSE = mean(mse_nls, na.rm = TRUE)),
                   data.frame(Contam = alpha, Method = "MM",
                              MSE = mean(mse_mm)))
}

ggplot(mse_tbl, aes(Contam, MSE, colour = Method)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_log10() +
  scale_colour_manual(values = c(NLS = "grey", MM = "black")) +
  theme_classic() +
  labs(x = "Contamination fraction",
       y = "Mean squared error (log10 scale)")
```

# Uncertainty Comparison {#sec-uncertainty-quant}

```{r}
#| label: uncertainty-comparison
#| eval: false

test_ages <- c(500, 1000, 1500)
results <- matrix(NA, nrow = length(test_ages), ncol = 4)
colnames(results) <- c("Age", "Pooled_Width", "Mixed_Width", "Width_Ratio")

# pooled NLS pieces
cf        <- coef(nls_pooled)
V_nls     <- vcov(nls_pooled)
sig2_nls  <- deviance(nls_pooled) / df.residual(nls_pooled)
df_nls    <- df.residual(nls_pooled)

# NLME pieces 
beta       <- fixef(nlme_fit)
Vb         <- vcov(nlme_fit)
tau2       <- as.numeric(VarCorr(nlme_fit)["Asym", "Variance"])  # var(b_Asym)
sig2_nlme  <- nlme_fit$sigma^2

# logistic mean and gradient
f <- function(t, Asym, xmid, scal) Asym / (1 + exp((xmid - t)/scal))
grad <- function(t, Asym, xmid, scal) {
  z <- (xmid - t)/scal
  g <- 1 / (1 + exp(z))
  c(
    g,                              # d/dAsym
    -(Asym/scal) * g * (1 - g),    # d/dxmid
    (Asym * z / scal) * g * (1 - g) # d/dscal
  )
}

for (i in seq_along(test_ages)) {
  t <- test_ages[i]

  g_nls <- grad(t, cf["Asym"], cf["xmid"], cf["scal"])
  se_pred_nls <- sqrt(as.numeric(t(g_nls) %*% V_nls %*% g_nls) + sig2_nls)
  w_nls <- 2 * qt(0.975, df_nls) * se_pred_nls

  g_mx  <- grad(t, beta["Asym"], beta["xmid"], beta["scal"])
  var_mean <- as.numeric(t(g_mx) %*% Vb %*% g_mx)
  var_re   <- (g_mx[1]^2) * tau2
  se_pred_nlme <- sqrt(var_mean + var_re + sig2_nlme)
  w_nlme <- 2 * qnorm(0.975) * se_pred_nlme

  results[i, ] <- c(t, w_nls, w_nlme, w_nlme / w_nls)
}

print(round(results, 2))

```